---
title: "Introduction to owprm - Occasions within Phases Repeated Measures"
author: "W Gregory Alvord"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
library(nlme)
library(ggplot2)
```


```{r, echo = FALSE}
## Read in ascorbic data set
## load owprm package
library(owprm)
data(ascorbic)
## summary(ascorbic)
## str(ascorbic)
##
```

### 1. Introduction

_Repeated measures_ data arise in a wide array of different research environments.  In this context, ... 

>  The term 'repeated' is used here to describe measurements  which are made on the same characteristic on the same observational unit but on more than one occasion. (Crowder and Hand, 1990, p. 1)

Because the measurements for a particular individual, subject, patient or observational unit  are _repeated_, they are not _independent_ within the individual.  Therefore, special care must be taken in the statistical analysis of such data.  


We analyze some _repeated measures_ data taken from the textbook by Crowder and Hand (1990), Example 3.3 on page 32.  Twelve hospital patients underwent a dietary regime treatment.  Over the course of seven occasions, measurements were taken on ascorbic acid for each patient. The seven occasions were divided into three treatment phases - twice before, thrice during, and twice after the treatment regime.  The seven occasions occurred at weeks 1, 2, 6, 10, 14, 15, & 16. The ascorbic acid measurements are made on the _same_ patient (subject) over time, and therefore, are not statistically _independent_ within the individual patient. 

### 2. The Model

For this problem, the model may be written as

$$
y_{ijk} = \pi_{i} + \Phi_{j} + o_{k(j)} + \pi\Phi_{ij} + \pi o_{ik(j)}
$$


where $y_{ijk}$ is $ijk^{th}$ response for the $i^{th}$ patient on the $k^{th}$ occasion in the $j^{th}$ phase,  $\pi_{i}$ is the $i^{th}$ patient, $\Phi_{j}$ is the $j^{th}$ phase, $o_{k(j)}$ is the $k^{th}$ occasion _nested_ within the $j^{th}$ phase, and $\pi o_{ik(j)}$ is the _interaction_ between the $i^{th}$ patient and the $k^{th}$ occasion _nested_ within the $j^{th}$ phase.  In this analysis, the patients (`patient`), occasions (`occ`) and phases (`Phase`) are considered as _factors_ in the statistical setup.  In particular, the `occ` variable is a _factor_ variable, not a _numeric_ variable.   Therefore, this statistical model is _fully-saturated_.  It contains no error term. That is, there is no $\epsilon_{ijk}$ represented in the model.  The error term for testing the hypothesis of interest in the `anova` is constructed in a special way, as described in the sequel.  

The patient, `patient`, represented as $\pi_{i}$, is a _random_ factor.  A `patient` is _randomly_ selected from a potentially large population of patients.  However,  `Phase` (Phase) and `occ` (Occasion) are _fixed_ factors.  They are measured or controlled by the investigator and are comprised of a relatively low number of finite levels.  In the model, any interaction between a fixed factor and a random factor is a _random_ factor. 

In order to test for the hypothesis of whether some relative upward or downward _shift_ occurs among `Phase`'s, it is necessary to construct the $F$ statistic in a non-traditional way.  To obtain the correct $F$ statistic, it is necessary to generate the _expected mean-squares_ from the structural model.  The _expected mean-squares_ can be derived from different algorithms, such as the _Cornfield-Tukey_ algorithm, as outlined, for example, in Winer (1971).       
      

### 3. Plots and Data Description

Figure 1 shows a plot of the data from Crowder and Hand (1990). This plot is constructed from a `groupedData` data frame from package `nlme` (Pinheiro \& Bates, 2000).  The data appear to show a rise at week 6 and a fall (drop) at week 15.  

```{r, echo = FALSE}
## names(ascorbic)
## with(ascorbic, levels(Phase))
```


```{r, echo = FALSE, fig.width = 5, fig.height = 3.5, fig.cap = "Reaction of twelve `patient`s to treatment (`Rx`) across seven Occasions (`occ`) in three `Phase`s."}
##
ascorbic.gD.df <-
groupedData(ascorbic.acid ~ week | patient,
data = ascorbic)
## plot ascorbic.gD.df
plot(ascorbic.gD.df, outer = ~ 1, key = FALSE,
ylab = 'Ascorbic Acid', xlab = 'Week', aspect = 0.4, main = 'Reaction of Patients to Rx')
##
```


Figure 2 shows an alternative plot of the `ascorbic` data, emphasizing that the occasions on the $x$-axis are to be considered as a _factor_ dimension as opposed to a _numeric_ dimension.   This plot is constructed from the `ggplot2` package (Wickham, 2009).  


```{r, echo = FALSE, fig.height = 4.5, fig.width = 6.5, fig.cap = "Reaction of twelve `patient`s to treatment (`Rx`) across seven Occasions (`occ`) in three `Phase`s."}
##
## ***** Get ggplot
##
## Create ggplot2 plot object
library(ggplot2)
##
occ.p0 <- ggplot(ascorbic, aes(occ, ascorbic.acid)) ##
set.seed(11) ## for jittered points to stay fixed
occ.p1 <- occ.p0 +
geom_boxplot(stat = 'boxplot', outlier.shape = 3) + labs(y = 'Ascorbic Acid', x = 'Occasion',
title = 'Ascorbic Acid Responses for 12 Patients
Across 7 Occasions in 3 Phases')
## occ.p1 ## basic boxplot
##
set.seed(11) ## for jittered points to stay fixed
occ.p2 <- occ.p1 +
geom_point(position = position_jitter(width = 0.2), aes(colour = patient))
## occ.p2 # add points color-coded by patient
##
set.seed(11) ## for jittered points to stay fixed
occ.p3 <- occ.p2 +
geom_vline(xintercept = c(2.5, 5.5), col = 'blue', lwd = 1.2, linetype = 'longdash')
## occ.p3 # add vertical lines to separate Phases
##
set.seed(11) ## for jittered points to stay fixed
occ.p4 <- occ.p3 +
annotate('text', x = 1.5, y = 1.6, label = 'pre Rx') +
annotate('text', x = 4, y = 0.3, label = 'Rx') +
annotate('text', x = 4, y = 0.2, label = '(treatment)') +
annotate('text', x = 6.5, y = 1.6, label = 'post Rx')
## occ.p4 # label Phases
##
set.seed(11) ## for jittered points to stay fixed
occ.p5 <- occ.p4 +
annotate('text', x = 1.5, y = 1.7, label = 'Phase 1') +
annotate('text', x = 4, y = 0.4, label = 'Phase 2') +
annotate('text', x = 6.5, y = 1.7, label = 'Phase 3')
occ.p5 # further label Phases
##
## ***** End ggplot
##
```


On the Occasions axis, the responses  are _jittered_ to avoid over-plotting.  Legends for the patients, `patient`, are included in the plot.  For these data, we have three phases, `Phase`, which are designated as `pre`, `Rx`, and `post` to correspond to the respones _before_, _during_, and _after_ the treatment regimen.  With respect to the spread or _variability_ of the data, it appears that the ascorbic acid responses are reasonably homogeneous throughout the seven occasions.  

In these analyses, we are interested in detecting `Phase` shifts.  The responses appear to be relatively low in the first phase, `pre`.  They are followed by an increase in the second phase, `Rx` and then by a drop in the third phase, `post`.  



### 4. Analysis


In this problem, there is **_one_** hypothesis of interest.  It is motivated by the question: Is there a difference in the mean levels of responses across the three `Phase`s under consideration, `pre`, `Rx`, and `post`? The criterion variable is ascorbic acid, i.e.,  `ascorbic.acid` in the `ascorbic` data frame.  Here is an abbreviated `anova` (analysis of variance) summary table for this analysis. 

```{r, echo = FALSE}
##
aovObj <- aov(ascorbic.acid ~ patient*Phase*occ, data = ascorbic)
## (alternatively) aovObj <- aov(ascorbic.acid ~ patient + Phase + Phase:occ + patient:Phase + patient:Phase:occ, data = ascorbic)
## aovObj
## To avoid having warning messages sent to console
## suppressWarnings(owprm(aovObj))
owprm_aovObj <- suppressWarnings(owprm(aovObj))
owprm_aovObj$'Summary Table of aov object'
## summary(owprm_aovObj)
## display p value in prettier form
## options(digits = 4, scipen = 2)
## owprm_aovObj
## summary(owprm_aovObj)
##
```


    
The correct test for determining whether there is an effect due to `Phase` is constructed by taking the mean square for `Phase` (2.9805) and dividing it by the mean square for `patient:Phase` (0.1090), and entering the $F$ distribution with the appropriate degrees of freedom, shown here.  The printing of the probability value to four  significant digits is deliberate. 

```{r, echo = FALSE}
##
options(digits = 4, scipen = 2)
## owprm_aovObj
summary(owprm_aovObj)
##
```


```{r, echo = FALSE}
## Reset options to original digits and scipen
options(digits = 7, scipen = 0)
##
```


### 5. The `owprm` Package 

We examine/analyze this problem with the use of  the `owprm` package. First load the `owprm` package. 

```{r, echo = TRUE}
library(owprm)
data(ascorbic)
str(ascorbic)
```

`ascorbic.acid` is the criterion variable of interest.  We fit a _fully-saturated_ analysis of variance (`aov`) object with the interaction of three `factor` variables, `patient`, `Phase` and `occ`. Next we display the `aovObj`  

```{r, echo = TRUE}
aovObj <- aov(ascorbic.acid ~ patient*Phase*occ, data = ascorbic)
aovObj
```

Before proceeding further, use the `summary` method on the `aovObj`.    

```{r, echo = TRUE}
summary(aovObj)
```


We next apply the `owprm()` function to the `aovObj`.  Because the `aovObj` is fully saturated, a warning is printed to the screen.

```{r, echo = TRUE}
owprm_aovObj <- owprm(aovObj)
```

The warning messages can be avoided by executing the `suppressWarnings()` command. Now display the `owprm_aovObj` object.   

```{r, echo = TRUE}
owprm_aovObj <- suppressWarnings(owprm(aovObj))
owprm_aovObj
```

The `owprm_aovObj` object is a list containing two components: (1) `Anova Summary Table for Occ w/in Phases Rep Meas Analy`, and (2) `Summary of Saturated ANOVA Object`. The `Anova Summary Table for Occ w/in Phases Rep Meas Analy` shows the correct $F$ statistic under the null hypothesis of no `Phase` shifts, the numerator and denominator degrees of freedom, and the probability value.  The `Anova Summary Table for Occ w/in Phases Rep Meas Analy` simply shows the `anova` summary table for the fully saturated `aov` fitted object.  No probability values are shown because there is no "error" term in the fully saturated model. 

In practice, we are generally most interest in the first `Anova Summary Table for Occ w/in Phases Rep Meas Analy` component.  This conveniently provided by the `summary` method applied to the `owprm_aovObj` object.    

```{r, echo = TRUE}
summary(owprm_aovObj)
```

This result may be displayed in a "prettier" form with the following code.

```{r, echo = TRUE}
options(digits = 4, scipen = 2)
summary(owprm_aovObj)
```



### References

Alvord WG and Carchedi N (2015) 'Occasions Within Phases Repeated Measures Analysis with Derivation of Expected Mean Squares by the Cornfield-Tukey Algorithm', https://rpubs.com/ga4247/53980




Cornfield J and Tukey JW (1956) 'Average values of mean squares in factorials' _Annals of Mathematical Statistics_, **27**, 907-949. 

Crowder MJ and Hand DJ (1990) _Analysis of Repeated Measures_, Chapman and Hall, London. 

Del Prete GQ, ..., Alvord WG, ... Lifson JD (2014) 'Effect of SAHA administration on the residual virus pool in a model of combination antiretroviral therapy-mediated suppression in SIVmac239-infected Indian rhesus macaques.' _Antimicrobial Agents and Chemotherapy_, **58**(11), 6790-806.

Pinheiro jC and Bates DM (2000) _Mixed-Effects Models in S and S-Plus_, Springer, New York.

Pinheiro J, Bates D, DebRoy S, Sarkar D and R
Core Team (2015). _nlme: Linear and Nonlinear
Mixed Effects Models_. R package version 3.1-122,
<URL: http://CRAN.R-project.org/package=nlme>.


Winer BJ (1971) _Statistical Principles in Experimental Design_, 2nd ed., McGraw-Hill. 





